---
title: "HW2"
author: "Haojie Liu"
date: "2025-04-16"
output: pdf_document
---

### 1. Download the file country_demographics.csv from the Module 2 datasets folder. I want you to tell me something about the linear relationship between GNP (gross national product = crude measure of a country’s wealth) and LEXPF (the life expectancy of females). You will encounter some problems. There are two helpful hints below.

```{r}
c_demog <- read.csv("country_demographics.csv", sep=",", header=T)
c_demog2 <-c_demog[complete.cases(c_demog),]
```


#### A. Generate the correlation coefficient between GNP and female life expectancy. Interpret the value of the correlation coefficient.

```{r}
with(c_demog, cor(GNP, LEXPF))
```

#### B. (The NA GNP values might cause problems - use Hint 1 and continue.)

```{r}
with(c_demog2, cor(GNP, LEXPF))
```
GNP and female life expectancy are positively correlated, where they increase and decrese at the same time.


#### . Plot the relationship between GNP and female life expectancy. What do you see?

```{r}
with(c_demog2, plot(GNP, LEXPF))
```

Not clear relationship showing on the plot


#### D. (The GNP data exhibit considerable skew and this is impacting the relationship between the two variables. Get rid of the skew by logging GNP – use Hint 2 and continue)


```{r}
logGNP <- log(c_demog2$GNP)
with(c_demog2, plot(logGNP, LEXPF))
```

#### E. Take the logarithm of GNP and recalculate the correlation coefficient between logged GNP and female life expectancy. Compare the result with what you found above. Check the data again in another plot. Does the plot look less impacted by skew?

```{r}
with(c_demog2, cor(logGNP, LEXPF))
```

Positive relationship become stronger after transform the data.

Hint 1

> c_demog <- read.csv("country_demographics.csv", sep=",", header=T)

> c_demog2 <-c_demog[complete.cases(c_demog),]

Hint 2

> logGNP <- log(c_demog2$GNP)

 

### 2. Find two of your own shapefiles, one with at least 100 points and the other containing at least 50 areas (polygons). These could come from the same database. (Don’t use the shapefefiles on the class website.) A shapefile is a collection of data and information used to store locational and geographic information. When downloading a shapefile, be sure to obtain and include all of the necessary components of the shapfile. The ‘st_read’ command will not import a shapefile, unless all of the necessary components are present in the shapefile folder.

```{r}
library(tmap)
library(tmaptools)
library(raster)
library(sf)
library(dplyr)

LAschoolBound <- st_read("/Users/liuhaojie/Desktop/STATSM222/Week3/LAschoolbound")

SCUpoint <- st_read("/Users/liuhaojie/Desktop/STATSM222/Week3/SCUpoint")
```

#### A. Map one of the variables in the polygon shapefile and briefly interpret what it tells you.

```{r}
qtm(LAschoolBound)
```

A simple map of LA districts.

#### B. Map the points in the second shapefile, find the mean center and standard distance and interpret what you see.

```{r}

SCUpoint_wgs <- st_transform(SCUpoint, 4326)


mc_coords <- st_coordinates(SCUpoint_wgs) |> colMeans()

mc <- st_sfc(st_point(mc_coords), crs = 4326) |> st_transform(st_crs(SCUpoint))
mc <- st_sf(geometry = mc)

coords_proj <- st_coordinates(SCUpoint)
center_proj <- st_coordinates(mc)
dists <- sqrt((coords_proj[,1] - center_proj[1,1])^2 + (coords_proj[,2] - center_proj[1,2])^2)
sd_radius <- sd(dists)
sd_circle <- st_buffer(mc, dist = sd_radius)

tm_shape(LAschoolBound) + tm_polygons() + 
  tm_shape(SCUpoint) + 
  tm_dots(size = 0.1) + 
  tm_shape(mc) + 
  tm_symbols(shape = 21, col = "red", size = 0.5, border.col = "black")+ 
  tm_shape(sd_circle) + 
  tm_borders(lwd = 2, col = "blue")


```

We can see most of the college, university and schools are locate around downtown LA.

### 3. On the Module 2 Datasets page are two files – us_state_popln_history.csv and us_state_centroids.csv. (If the state centroids are not working for you, you might have to find your own coordinates (lon, lat) for the centroid of each state.) The population-history file reports the population of all 50 states and Washington DC over 10 year intervals from 1870 to 2020. You should filter out (delete) the observations for Alaska and Hawaii, focusing only on the contiguous US states. (You will have to deal with two NA “values” early on for Oklahoma. Changing those to 0 is a solution in this case.)


#### A. Read into R the state population data.

```{r}
statepop <- read.csv("us_state_popln_history.csv")
statepop[is.na(statepop)] <- 0
statepop <- statepop %>%
  filter(!state %in% c("Alaska", "Hawaii"))

centroids <- read.csv("us_state_centroids.csv")
pop_centroids <- centroids %>%
  inner_join(statepop, by = "state") %>%
  select(state, lon, lat, pop1960)
```


#### B. Find a shapefile that you can use to link to the state population totals.

```{r}
state_bound <- st_read("cb_2018_us_state_5m")

state_bound <- state_bound %>%
  filter(STUSPS %in% c(
    "AL", "AR", "AZ", "CA", "CO", "CT", "DC", "DE", "FL", "GA", 
    "IA", "ID", "IL", "IN", "KS", "KY", "LA", "MA", "MD", "ME", 
    "MI", "MN", "MO", "MS", "MT", "NC", "ND", "NE", "NH", "NJ", 
    "NM", "NV", "NY", "OH", "OK", "OR", "PA", "RI", "SC", "SD", 
    "TN", "TX", "UT", "VA", "VT", "WA", "WI", "WV", "WY"
  ))

pop_sf <- st_as_sf(pop_centroids, coords = c("lon", "lat"), crs = 4326)

# Plot over contiguous US map
tm_shape(state_bound) + 
  tm_borders() +
  tm_shape(pop_sf) + 
  tm_dots(size = "pop1960", col = "red", scale = 0.5, title.size = "Pop 1960") +
  tm_layout(title = "1960 State Population (Contiguous US)", legend.outside = TRUE)
```



#### C. Calculate the weighted mean center of the US population for each decade and show how it has moved on your tmap map. (You can calculate the weighted mean center in R just using the state centroids, or whatever other coordinates that you obtain, along with the population values.  Let us not worry about the transformation from lat-lon coordinates to an XY space for the moment.) Briefly report what you see.

```{r}
data <- inner_join(statepop, centroids, by = "state")

get_weighted_center <- function(df, lon_col, lat_col, weight_col) {
  x <- sum(df[[lon_col]] * df[[weight_col]]) / sum(df[[weight_col]])
  y <- sum(df[[lat_col]] * df[[weight_col]]) / sum(df[[weight_col]])
  return(c(lon = x, lat = y))
}

decades <- grep("^pop", names(data), value = TRUE)

centers <- lapply(decades, function(decade) {
  coords <- get_weighted_center(data, "lon", "lat", decade)
  data.frame(decade = gsub("pop", "", decade), lon = coords[1], lat = coords[2])
})

centers_df <- do.call(rbind, centers)

centers_sf <- st_as_sf(centers_df, coords = c("lon", "lat"), crs = 4326)

tm_shape(state_bound) +
  tm_borders() +
  tm_shape(centers_sf) +
  tm_symbols(shape = 21, col = "red", size = 0.4, border.col = "black") +
  tm_text("decade", size = 0.5, ymod = -0.5) +
  tm_layout(title = "Movement of US Population Weighted Mean Center")

```

The populaton center are moving towards to the west from 1870 to 2020.





